\documentclass{acm_proc_article-sp}
\usepackage{cite}
\usepackage{hyperref}
\sloppy

\begin{document}
\toappear{This is an unpublished draft. Please don't distribute without our permission.}
\title{Octokey: Public key authentication for the web}
\numberofauthors{2}
\author{
    \alignauthor Martin Kleppmann \\ \email{martin@kleppmann.com}
    \alignauthor Conrad Irwin \\ \email{conrad.irwin@gmail.com}}
\maketitle

\begin{abstract}
We propose Octokey, a cryptographic protocol for authenticating users, designed as a viable
replacement for password authentication in consumer internet use cases. When a user signs up to a
service, they provide a public key. Clients subsequently authenticate by presenting a challenge
signed with the user's private key, similar to SSH public key authentication. We also describe
protocols which allow users to log in to their accounts on multiple devices, provision new devices,
and revoke keys on devices that are lost or stolen.
\end{abstract}

\bigskip

\emph{Note: We, the authors, have a background in developing commercial web applications. In the
field of crypography, we are merely keen amateurs. We are aware that amateur cryptography is prone
to catastrophic flaws, and we fully expect such flaws to be found in the ideas we are proposing.
However, we have experience designing real-world systems with good user experiences, a perspective
that we feel is often lacking in security research. We hope that this paper can stimulate a
discussion that brings together the perspectives from research and industry, and we would like to
solicit constructive feedback from all sides.}

\input{intro}
\input{protocol}
\input{provision}
\input{mobile}

% TODO: OAuth-like thing by adding a third party's public key to your account

\bibliography{references}{}
\bibliographystyle{plain}
\end{document}
